# RStudio Exercise 2: Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

 ```{r setup, include=FALSE}
 library(knitr)
 opts_chunk$set(results = 'asis',      # This is essential (can also be set at the chunk-level)
                comment = NA, 
                prompt = FALSE, 
                cache = FALSE)

 library(summarytools)
 st_options(plain.ascii = FALSE,       # This is very handy in all Rmd documents
            #style = "rmarkdown"        # This too
            footnote = NA,             # Avoids footnotes which would clutter the results
            subtitle.emphasis = FALSE  # This is a setting to experiment with - according to
 )                                     # the theme used, it might improve the headings' 
                                       # layout
 ```

 ```{r, echo=FALSE}
 st_css()                              # This is a must; without it, expect odd layout,
 ```

Read data from the website:

```{r}
learning2014 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)

```

Summary statistics:

```{r, echo=FALSE, message=FALSE}

dfSummary(learning2014, plain.ascii = FALSE, style = "grid", 
          graph.magnif = 0.75, valid.col = FALSE, na.col = FALSE, tmp.img.dir = "/tmp")
```